---
title: "Cyclist Data Trips"
author: "Marco Antonio Pereira Esteves"
date: "2025-02-20"
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE, warning=FALSE}
# install.packages("tidyverse")
library(tidyverse)

# install.packages("data.table")
library(data.table)

library(conflicted)

conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("wday", "lubridate")
```

### Reading the .csv files and merging into one data frame:

```{r message=FALSE, warning=FALSE}
file_list <- list.files(pattern = "*.csv")

trips_2024_df <- rbindlist(lapply(file_list, fread))

head(trips_2024_df)
```

### Remove lat, long fields
```{r message=FALSE, warning=FALSE}
trips_2024_df <- trips_2024_df %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

### Add columns that list the date, month, day, and year of each ride

```{r}
trips_2024_df$date <- as.Date(trips_2024_df$started_at)
trips_2024_df$month <- format(as.Date(trips_2024_df$date), "%m")
trips_2024_df$day <- format(as.Date(trips_2024_df$date), "%d")
trips_2024_df$year <- format(as.Date(trips_2024_df$date), "%Y")
trips_2024_df$day_of_week <- format(as.Date(trips_2024_df$date), "%A")
```

### Add a "ride_length" calculations (in seconds)
```{r}
trips_2024_df$ride_length <- difftime(trips_2024_df$ended_at,trips_2024_df$started_at)
```

# Convert "ride_length" to numeric so we can run calculations on the data
```{r}
trips_2024_df$ride_length <- as.numeric(as.character(trips_2024_df$ride_length))
is.numeric(trips_2024_df$ride_length)
```

### Cleaning bad data:

```{r}
trips_2024_df_clean <- trips_2024_df[!(trips_2024_df$start_station_name == "HQ QR" | trips_2024_df$ride_length<1),]
```

##### The total entries went from 5.860.568 to 5.858.968


### Performing calculations with the ride length column:

#### Ride Length Mean

```{r}
mean_trips <- mean(trips_2024_df_clean$ride_length)

mean_trips_hms <- round(seconds_to_period(mean_trips))

print(mean_trips_hms)
```

### Median Ride Length

```{r}
median_trips <- median(trips_2024_df_clean$ride_length)

median_trips_hms <- round(seconds_to_period(median_trips))

print(median_trips_hms)
```


#### Max Ride Length

```{r}
max_trips <- max(trips_2024_df_clean$ride_length)

max_trips_hms <- round(seconds_to_period(max_trips))

print(max_trips_hms)
```

#### Min Ride Length

```{r}
min_trips <- min(trips_2024_df_clean$ride_length)

min_trips_hms <- round(seconds_to_period(min_trips))

print(min_trips_hms)
```


#### Most ocurring day of week

```{r}
freq_table <- table(trips_2024_df_clean$day_of_week)
modes <- names(freq_table[freq_table == max(freq_table)])
print(modes)
```

#### Least ocurring day of week

```{r}
freq_table <- table(trips_2024_df_clean$day_of_week)
modes <- names(freq_table[freq_table == min(freq_table)])
print(modes)
```

#### Average ride length for casual and member riders

```{r}
total_avg_ride_duration <- trips_2024_df_clean %>%
  group_by(member_casual) %>%
  summarise(avg_seconds = mean(ride_length, na.rm = TRUE)) %>%
  mutate(avg_time = seconds_to_period(round(avg_seconds))) %>%
  select(member_casual, avg_seconds)
print(total_avg_ride_duration)
```

#### Average ride length for users by day of the week

```{r}
trips_2024_df_clean %>%
  group_by(day_of_week) %>%
  summarise(avg_seconds = mean(ride_length, na.rm = TRUE)) %>%
  mutate(avg_time = seconds_to_period(round(avg_seconds))) %>%
  select(day_of_week, avg_time)
```

#### Total rides for users by day of the week

```{r message=FALSE, warning=FALSE}
total_rides_by_day <- trips_2024_df_clean %>%
  group_by(day_of_week, member_casual) %>%
  summarise(total_rides = n())

print(total_rides_by_day)
```

#### Compare members and casual users
```{r}
aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual, FUN = mean)
aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual, FUN = median)
aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual, FUN = max)
aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual, FUN = min)
```


#### See the average ride time by each day for members vs casual users
```{r}
trips_2024_df_clean$day_of_week <- ordered(trips_2024_df_clean$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual + trips_2024_df_clean$day_of_week, FUN = mean)
```

#### Analyze ridership data by type and weekday
```{r message=FALSE, warning=FALSE}
trips_2024_df_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n()							 
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)
```


#### Visualization for the number of rides by rider type

```{r message=FALSE, warning=FALSE}
trips_2024_df_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of Rides by User Type and Day of the Week",
    subtitle = "Comparing casual and member riders across weekdays",
    x = "Day of the Week",
    y = "Number of Rides",
    fill = "User Type"
  )  +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```


#### Visualization for average duration

```{r message=FALSE, warning=FALSE}

trips_2024_df_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Duration by User Type and Day of the Week",
    subtitle = "Comparing casual and member riders across weekdays",
    x = "Day of the Week",
    y = "Average Duration (in seconds)",
    fill = "User Type"
  )  +
  theme_minimal()
```

#### Visualization for total average duration

```{r}

trips_2024_df_clean %>%
  group_by(member_casual) %>%
  summarise(avg_seconds = mean(ride_length, na.rm = TRUE)) %>%
  mutate(avg_time = seconds_to_period(round(avg_seconds))) %>%
  select(member_casual, avg_time)  %>% 
  ggplot(aes(x = member_casual, y = avg_time, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Total Average Duration by User Type",
    subtitle = "Comparing casual and member riders",
    x = "Member Type",
    y = "Average Duration",
    fill = "User Type"
  )  +
  theme_minimal()
```


#### Export
```{r}
avg_ride_length <- aggregate(trips_2024_df_clean$ride_length ~ trips_2024_df_clean$member_casual + trips_2024_df_clean$day_of_week, FUN = mean)
write.csv(avg_ride_length, file = 'avg_ride_length.csv')

write.csv(total_rides_by_day, file = 'total_rides_users_wday.csv')

write.csv(total_avg_ride_duration, file = 'total_avg_ride_duration.csv')

```